---
title: "Getting Started with Pytekt in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Pytekt in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction to Pytekt for R

The `pytekt` R package provides native C++ numerical acceleration for common vector operations, reductions, matrix operations, and streaming algorithms, as well as an optional seamless bridge to Pytekt's full Python machine learning and scientific suite via `reticulate`.

## 1. Native High-Performance Numerical Kernels

All native kernels in `pytekt` are compiled via Rcpp using high-performance C++14 routines.

```{r setup}
library(pytekt)

# Sample dataset
x <- c(10.0, 25.5, -4.2, 18.0, 99.1, 3.7)
```

### 1D Reductions & Summary Statistics

```{r reductions}
# Fast sum and mean
pytekt_sum(x)
pytekt_mean(x)

# Variance and standard deviation
pytekt_variance(x, ddof = 1)
pytekt_std(x, ddof = 1)

# Extrema and indices (1-based for R)
pytekt_min(x)
pytekt_max(x)
pytekt_argmin(x)
pytekt_argmax(x)
```

### Vector Distance & Similarity

```{r distance}
a <- c(1.0, 2.0, 3.0)
b <- c(4.0, 5.0, 6.0)

# Euclidean Distance & Cosine Similarity
pytekt_euclidean_distance(a, b)
pytekt_cosine_similarity(a, b)
```

### Streaming & Window Operations

```{r streaming}
# Cumulative Prefix Sum
pytekt_prefix_sum(x)

# Rolling window statistics
pytekt_rolling_mean(x, window = 3)
pytekt_rolling_sum(x, window = 3)

# Streaming histogram
pytekt_histogram(x, bins = 5, lo = -10.0, hi = 100.0)

# Block / Chunk Statistics
pytekt_chunk_statistics(x, chunk_size = 2)
```

### Fast Matrix Multiplication

```{r matmul}
# 2x2 matrices flattened row-major
A <- c(1, 2, 3, 4)
B <- c(5, 6, 7, 8)

pytekt_matrix_multiply(A, B, m = 2, n = 2, p = 2)
```

## 2. Using Python Machine Learning & Scientific Models in R

When `reticulate` and the Python `aqwel-aion` package are installed, you can access the complete ML, physics, and astronomy models directly from R:

```R
if (pytekt_available()) {
  pytekt <- import_pytekt()
  
  # Load built-in research benchmark dataset
  iris_ds <- pytekt$datasets$load_iris()
  
  # Train a Gaussian Naive Bayes classifier
  scaler <- pytekt$preprocessing$StandardScaler()
  X_scaled <- scaler$fit_transform(iris_ds$data)
  
  clf <- pytekt$models$GaussianNB()
  clf$fit(X_scaled, iris_ds$target)
  
  preds <- clf$predict(X_scaled)
  acc <- pytekt$metrics$accuracy_score(iris_ds$target, preds)
  cat("Model Accuracy:", acc, "\n")
}
```
